Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2021-11-07 docs/00-manuscripts/mdpi/manuscript.tex
numerous minor editorial changes

	%\setlength{\itemsep}{-\parsep}
	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborne laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FFS}{forward feature selection}
	\acro{FPR}{false positive rate}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic i Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatter plot smoothing}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NDMI}{normalized difference moisture index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{OLS}{ordinary least squares}
	\acro{OMNBR}{optimized multiple narrow-band reflectance}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PISR}{potential incoming solar radiation}
	\acro{PLS}{partial least-squares}
	\acro{POV}{proportion of variance explained}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential-based model optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
	\acro{PC}{principal component}
\end{acronym}

@@

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Besides the availability of code and manuscript sources, a static webpage is available at (\url{https://pat-s.github.io/2019-feature-selection}), listing more side-analyses that were carried out during the creation of this study.

@@

\subsection{PCA-based variance analysis of feature sets}

@@

% learner
% - SVM Car is the overall winner
% - Lasso quite stable (varies within 1 - 4), ridge a bit more variable (2 - 6)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Relief} filter achieved the best overall performance (RMSE of 28.09 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalty showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Difference to other feature sets were very small for lasso (below one p.p.) and a bit higher for ridge (between two and five p.p. (VI)).
XGBoost shows bad performances for some feature sets and fills the ten last places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
Using filters for RF showed a substantial increase in performance for all tasks with the exception of NRI, for which the difference among all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost showed a high dependency on filtering the data: In two out of six tasks using no filter resulted in the worst or second worst performance.
XGBoost shows the highest overall differences between filters for a single task: for feature set HR, the range is up to 15 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

@@

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter achieved in only one instance (RF on HR-NRI-VI) the best performance among the used filters (\autoref{fig:filter-effects-borda}).
For RF and XGBoost it most often ranked within the first 50\% with respect to all filters of a specific task.
For XGBoost on the NRI task, the Borda filter scored the second worst performance.

% Selected percentage of features across plots
Large differences were observed between the numbers of features selected during tuning for the subsequent fitting process across each learner and plot.
While for RF least features were selected when Luiando or Oiartzun were the test set ($n = 1$), more than 90\% of all features were used for test sets Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set, (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots among all learners, with 24\% for test set Laukiz2 being the highest single plot value.
In general every learner behaved quite differently for each plot and no overall pattern could be observed.

@@

% plot performance results
\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.7\textwidth] {performance-results-1.pdf}
		\caption{Predictive performance in RMSE (p.p.) of models across tasks. Different feature sets are shown on the y-axis. Labels show the feature selection method (e.g. NF = no filter, Car = 'Carscore', Info Gain = 'Information Gain', Borda = 'Borda').}\label{fig:perf-result}
	\end{center}
\end{figure}

@@

The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For dataset HR most features cluster around the infrared region (920 nm - 1000 nm) (six out of ten) while for VI eight out of ten concentrate on the wavelength range of 700 nm - 750 nm (the so called \enquote{red edge}).
For feature set HR, four features in the infrared region (920 nm - 1000 nm) were identified by the model to be most important (causing a mean decrease in RMSE of around 1 p.p.).
Overall, most features (excluding the top five respectively) showed only a small importance with average decreases in RMSE below 0.5 p.p..

>>>>

	%\setlength{\itemsep}{-\parsep}
	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborne laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FFS}{forward feature selection}
	\acro{FPR}{false-positive rate}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic i Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatterplot smoothing}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NDMI}{normalized difference moisture index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{OLS}{ordinary least squares}
	\acro{OMNBR}{optimized multiple narrow-band reflectance}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PISR}{potential incoming solar radiation}
	\acro{PLS}{partial least squares}
	\acro{POV}{proportion of variance explained}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential model-based optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true-positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
	\acro{PC}{principal component}
\end{acronym}

@@

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Apart from the availability of code and manuscript sources, a static webpage is available at (\url{https://pat-s.github.io/2019-feature-selection}), which includes additional side analyses that were carried out during the creation of this study.

@@

\subsection{Principal component analysis of feature sets}

@@

% learner
% - SVM Car is the overall winner
% - Lasso quite stable (varies within 1 - 4), ridge a bit more variable (2 - 6)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Relief} filter achieved the best overall performance (RMSE of 28.09 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalties showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Differences to other feature sets were very small for lasso (below one p.p.) and a bit larger for ridge regression (between two and five p.p.).
XGBoost showed very poor performances for some feature sets and fills the ten last places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
The use of filters with RF resulted in a substantial increase in performance in all tasks with the exception of NRI, in which the difference among all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost's performance depended strongly on feature selection: In two out of six tasks, using no filter resulted in the worst or second worst performance.
XGBoost showed the highest overall differences between filters for a single task: for feature set HR, the range is up to 15 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

@@

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter achieved in only one instance (RF on HR-NRI-VI) the best performance among the used filters (\autoref{fig:filter-effects-borda}).
For RF and XGBoost it most often ranked within the better half among all filters of a specific task.
For XGBoost on the NRI task, the Borda filter scored the second worst performance.

% Selected percentage of features across plots
The number of features selected based on model tuning strongly varied across learners and plots.
While for RF, the least features were selected when Luiando or Oiartzun was the test set ($n = 1$), more than 90\% of all features were used for test sets Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set, (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots and among all learners, the single highest fraction being 24\% for test set Laukiz2.
In general, all learners behaved rather differently on the various plots, and therefore no overall pattern could be identified.

@@

% plot performance results
\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.7\textwidth] {performance-results-1.pdf}
		\caption{Predictive performance in RMSE (p.p.) of models across tasks. Different feature sets are shown on the y-axis. Labels show the feature selection method (e.g., NF = no filter, Car = 'Carscore', Info Gain = 'Information Gain', Borda = 'Borda').}\label{fig:perf-result}
	\end{center}
\end{figure}

@@

The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For the HR dataset, most (i.e., six out of ten) relevant features clustered around the infrared region (920 nm - 1000 nm), while for VI, eight out of ten concentrate within the wavelength range of 700 nm - 750 nm (the so called \enquote{red edge}).
For HR, four features in the infrared region (920 nm - 1000 nm) were identified by the model to be most important, being assciated with a mean decrease in RMSE of around 1 p.p.
Overall, apart from the top five features, the vast majority of features showed only a small importance with average decreases in RMSE below 0.5 p.p.
